---
title: 'Part 2: Data Cleaning'
author: "Matt Steele"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
---

------------------------------------------------------------------------

## Resources

-   [Tidyverse Documentation](https://www.tidyverse.org/)

-   [O'Reilly Learning
    Platform](https://databases.lib.wvu.edu/connect/1540334373)

    -   R for Data Science, 2nd Edition

    -   R Programming for Statistics and Data Science
    
    -   Text Mining with R

------------------------------------------------------------------------

```{r}
#| label: working directory
#| include: false

setwd("C:/Users/Matt/Documents/GitHub/r_data_cleaning_fall23")


```

<br />

# Part 1: About Tidyverse

```{r}
#| label: install tidyverse
#| eval: false

install.packages("tidyverse")

```



```{r}
#| label: load tidyverse
#| eval: false

library(tidyverse)

```

Tidyverse is a collection of packages focused on data analysis and data
visualizations that share an underlying design philosophy, grammar, and
data structures.

|                                                     |                                                                   |
|-------------------------|-----------------------------------------------|
| [tibble](https://tibble.tidyverse.org/)             | lighter and more user-friendly version of data frames             |
| [tidyr](https://tidyr.tidyverse.org/)               | create tidy and meaningfully arranged data                        |
| [readr](https://readr.tidyverse.org/)               | better importation of data into R                                 |
| [ggplot](https://ggplot2.tidyverse.org/)            | data visualization functions                                      |
| [dplyr](https://dplyr.tidyverse.org/)               | data manipulation tools                                           |
| [lubridate](https://lubridate.tidyverse.org/)       | clean dates and times                                             |
| [purr](https://purrr.tidyverse.org/)                | better functional programming                                     |
| [forcats](https://forcats.tidyverse.org/index.html) | handle, clean, and manipulate categorical variables               |
| [haven](https://haven.tidyverse.org/)               | read and write data formats from proprietary statistical pacakges |

: Packages Includes in Tidyverse

<br />

------------------------------------------------------------------------

# Part 2: Loading Data with Tidyverse

<br />

---

## Loading Data

<br />

---

### Tibble function: read_csv

this function allows you to read a csv file into a tibble data frame

```{r}
#| label: read_csv function
#| eval: false

reviews_df <- read_csv("customer_reviews.csv")

```

<br />

------------------------------------------------------------------------

### Tibble function: view

tibble also provides a lighter way for you to view you entire data
frame.

```{r}
#| label: view function
#| eval: false

view(reviews_df)

```

<br />

------------------------------------------------------------------------

### The Pipe Operator

The pipe operator allows you to run commands or operation on a single
object based on an order of operations

-   let's say you want to see the **age**, **rating**, and **feedback**
    of customer who shopped in the *dresses* department

```{r}
#| label: pipe operator example
#| eval: false

view(reviews_df)

reviews_df |> # object we are working on
  filter(Department_Name == "Dresses") |> # first operation
  select(Clothing_ID, Age, Rating, Positive_Feedback_Count) # second operation

# order of operations matter

reviews_df |> # object we are working on
  select(Clothing_ID, Age, Rating, Positive_Feedback_Count) |> # first operation
  filter(Department_Name == "Dresses") # second operation

# why did this not work?


```

<br />

------------------------------------------------------------------------

### Readr function: read_excel \| read_xlsx

this function allows you to read Excel files in a tibble data frame


```{r}
#| label: read excel files
#| eval: false

breed_rank_all <- readxl::read_xlsx("dog_breeding.xlsx", sheet = 1)
breed_traits <- readxl::read_xlsx("dog_breeding.xlsx", sheet = 2)

```

<br />

------------------------------------------------------------------------

### Dplyr function: mutating join functions

this function allows you to join multiple data frames together by a
common variable(s)

|            |                                                       |
|------------|-------------------------------------------------------|
| left_join  | if observations in (y) are in (x) then add (y) to (x) |
| right_join | if observations in (x) are in (y) then add (x) to (y) |
| full_join  | add all observations from (x) to (y)                  |
| anti_join  | remove entries from data frame (y)                    |

: mutating join functions

<br />

```{r}
#| label: mutating joins
#| eval: false 


breed_rank_all # ranks of dog breeds 
breed_traits # traits of dog breeds

breed_df <- breed_rank_all |> 
  left_join(breed_traits, by = "Breed")

breed_df

```

<br />

---

### Haven package: Read non-proprietary data files

The package Haven allows you to read and export non-proprietary files
for SPSS, SAS, and STATA

```{r}
#| read spss - haven
#| eval: false


demographics_df <- haven::read_sav("demographics.sav")

```

<br />

#### Glimpse function

like the str() function in base r, this allow you see the structure of
your data but in a more compact manner

```{r}
#| label: glimpse function
#| eval: false

glimpse(demographics_df)

```

<br />

------------------------------------------------------------------------

# Part 3: Cleaning Data

|                                                                   |                                                                |
|-----------------------|-------------------------------------------------|
| [filter](https://dplyr.tidyverse.org/reference/filter.html)       | retains or filters out observations based on variable criteria |
| [select](https://dplyr.tidyverse.org/reference/select.html)       | retains or filters out variables                               |
| [arrange](https://dplyr.tidyverse.org/reference/arrange.html)     | sorts variables                                                |
| [mutate](https://dplyr.tidyverse.org/reference/mutate.html)       | change variable's observations OR create a new variable and observations using observations from another variable                       |
| [group_by](https://dplyr.tidyverse.org/reference/group_by.html)   | group observations                                             |
| [summarise](https://dplyr.tidyverse.org/reference/summarise.html) | get descriptive statistics about a variable                    |

: Main Tidyverse Functions

<br />

------------------------------------------------------------------------


## Dplyr function: filter

the filter function allows you to *filter* your data frame using a
observations in a variable

```{r}
#| label: filter function
#| eval: false

reviews_df

# let's filter the data frame so we are seeing people who bought pants

reviews_pants <- reviews_df |> 
  filter(Class_Name == "Pants")

reviews_pants

```

<br />

### Boolean operators

boolean operators allows you to build criteria in your code

|     |                       |
|-----|-----------------------|
| &   | AND                   |
| \|  | OR                    |
| ==  | EQUAL                 |
| !=  | NOT EQUAL             |
| \<  | LESS THAN             |
| \>  | GREATER THAN          |
| \<= | LESS THAN OR EQUAL    |
| \>= | GREATER THAN OR EQUAL |

: Boolean operators

<br />

```{r}
#| label: filter with more than one criteria
#| eval: false

# let's filter the data frame for people who did not buy pants and are over 50 years old

reviews_pants50 <- reviews_df |> 
  filter(Class_Name != "Pants" & Age > 50)

reviews_pants50

```

<br />

------------------------------------------------------------------------

## Dplyr function: select

the select function allows you to *keep* or *discard* variables

```{r}
#| label: select function
#| eval: false

# let's just keep the numeric information

reviews_select <- reviews_df |> 
  select(Clothing_ID, Age, Rating, Recommended_IND)

reviews_select

# let's remove the division and department name variables from our data frame

reviews_notSelect <- reviews_pants50 |> 
  select(-Division_Name, -Department_Name)

reviews_notSelect

```

<br />

------------------------------------------------------------------------

## Dplyr function: mutate

the mutate function allows you to change variables OR create new
variables based on the observations in the variables

*NOTE: if you name your variable as an existing variable, it WILL overwrite the existing varaible. If you give it a new name, it will create a new variable

```{r}
#| label: mutate function
#| eval: false

demographics_df

demographics_mutate <- demographics_df |> 
  mutate(income_new = income/1000) |> 
  relocate(income_new, .after = income)

demographics_mutate

# let's overwrite the old variable

demographics_overwrite <- demographics_df |> 
  mutate(income = income/1000)

demographics_overwrite

```

<br />

------------------------------------------------------------------------

## Dplyr function: arrange

the arrange function allows you to sort variables

```{r}
#| label: arrange function
#| eval: false

# let's look in the customers review data frame and see what customers gave the highest rating

reviews_df |> 
  arrange(desc(Rating))

# you can add multiple criteria to your sort

reviews_df |> 
  arrange(desc(Rating), Class_Name) |> 
  select(Clothing_ID, Rating, Class_Name)

```

<br />

---

## Dplyr function: group_by & summarise

the **group_by** function allows you to group common observations in a variable and **summarise** function allows you to get descriptive statistics about the groupings

```{r}
#| label: group_by and summarise
#| eval: false

reviews_group <- reviews_df |> 
  group_by(Class_Name) |> 
  summarise(mean(Rating)) # get the average for ratings

reviews_group

# you can add as many summaries as you want

reviews_group <- reviews_df |> 
  group_by(Class_Name) |> 
  summarise(n = n(), # get the instances count
            positive_tot = sum(Positive_Feedback_Count), # get the sum of positive feedback 
            positive_avg = mean(Positive_Feedback_Count), # get the average of positive feedback
            rating = mean(Rating)) # get the average rating


# you can also add multiple groups

reviews_group <- reviews_df |> 
  group_by(Class_Name, Recommended_IND) |> 
  summarise(n = n(), 
            positive_tot = sum(Positive_Feedback_Count), 
            positive_avg = (sum(Positive_Feedback_Count)/n), 
            rating = mean(Rating))

```



<br />

------------------------------------------------------------------------

## Base Function: as.character 

This function along with **mutate** will allow you to change the data type of a variable. For this example we are going to recode the *Clothing_ID* variable to interpret the data type as a *character* instead of a *double*

```{r}
#| label: data type recode
#| eval: false

reviews_df <- reviews_df |> 
  mutate(Clothing_ID = as.character(Clothing_ID))

reviews_df

```



## Forcats function: as_factor

This function allows you to redefine a variable value as a factor using the **mutate** function. For this example we are going redefine the recommend variable in the data frame we created when we grouped the customers.

```{r}
#| label: recode factors
#| eval: false

is.factor(reviews_group$Recommended_IND)

reviews_df <- reviews_df |> 
  mutate(Recommended_IND = as_factor(Recommended_IND))

reviews_df

```

<br>

## Dplyr function: mutate_at

You can also change the data type of multiple variables using the **mutate_at** function. For this example we will change the division, department, and class variables from *characters* into *factors*.

```{r}
#| label: recode multiple factors
#| eval: false

reviews_df <- reviews_df |> 
  mutate_at(vars(Division_Name, Department_Name, Class_Name), as_factor)

reviews_df

```


<br />

---

## Dyplr Function: recode

we can rename the values of observations within a variable using the **mutate** function in combination with the **recode** or **recode_factor** functions

```{r}
#| label: recode observations
#| eval: false

# we can look at what levels a factor variable has, so long as r is interpreting it as a factor variable



levels(reviews_df$Recommended_IND)

# let's change the values from 1/0 to yes/no

reviews_df <- reviews_df |> 
  mutate(Recommended_IND = recode_factor(Recommended_IND, 
                                         "1" = "Yes", 
                                         "0" = "No"))


reviews_recode

```

<br />

---


## Tidyr Function: Pivot wider

we can also do the inverse of pivot longer and widen our data

```{r}
#| label: pivot wider
#| eval: false


# let's create new variables from the recommend variable

reviews_wide <- reviews_df |> 
  group_by(Class_Name, Recommended_IND) |> 
  summarise(number = n(),
            feedback = mean(Positive_Feedback_Count)) |> 
  pivot_wider(names_from = Recommended_IND, values_from = c(number, feedback))

reviews_wide

```


<br />

---

## Dplyr function: rename

the rename function allows you rename variables in your data frame

```{r}
#| label: rename
#| eval: false

glimpse(reviews_wide)

reviews_wide <- reviews_wide |> 
  rename("Not_Recommend" = number_No, "Recommend" = number_Yes)

reviews_wide


```


<br />

---

## Tidyr function: drop_na

we can remove **all** missing data from data frames or variables using the drop_na function

```{r}
#| label: drop na
#| eval: false

reviews_wide

# we can see if are data frame has missing NA values using the is.na function.

is.na(reviews_wide)
which(is.na(reviews_wide))

# because there are missing values we cannot calculate some descriptive statistics

mean(reviews_wide$Not_Recommend)

# we can drop all NA values from the data frame

reviews_dropNA <- reviews_wide |> 
  drop_na()

reviews_dropNA

mean(reviews_dropNA$Not_Recommend)

# we can also just drop NAs from a variable

reviews_dropNA_var <- reviews_wide |> 
  drop_na(Not_Recommend)

reviews_dropNA_var

```


<br />

---

## Tidyr function: replace_na

you can also recode the NA values for observations with the replace_na function

```{r}
#| label: replace_na
#| eval: false

reviews_wide 

# let's replace the NAs in the Not recommended to 0 since there were no reviews given

reviews_replaceNA <- reviews_wide |> 
  mutate(Not_Recommend = replace_na(Not_Recommend, 0))

reviews_replaceNA

mean(reviews_replaceNA$Not_Recommend)

```

<br />

------------------------------------------------------------------------



```{r}
#| label: load covid data
#| eval: false

covid19 <- read_csv("time_series_covid19_confirmed_US.csv")

view(covid19)


```

<br />

---

## Tidyr function: Pivot Data

pivoting data allows you to either move variables into observations or
observations into variables.

```{r}
#| label: pivot longer
#| eval: false

# let's turn the date variables into observations

covid19_longer <- covid19 |> 
  pivot_longer(!Combined_Key, names_to = "Date", values_to = "Total")


covid19_longer

```

<br />



------------------------------------------------------------------------

## Tidyr function: Separate, Unite, and Gather

these functions allow you to split up or combine variables in your data
frame

```{r}
#| label: separate function
#| eval: false

covid19_separate <- covid19_longer |> 
  separate_wider_delim(Combined_Key, 
                       delim = ", ", 
                       names = c("City", "State", "Country"),
                       too_few = "align_start") |> 
  select(-Country)

covid19_separate

```

<br />

------------------------------------------------------------------------

## Lubridate function: as_date

the lubridate package allows you to clean and manipulate data variables

```{r}
#| label: change variable value to data
#| eval: false

glimpse(covid19_separate)

covid19_date <- covid19_separate |> 
  mutate(Date = lubridate::dmy(Date))

covid19_date

## we can even use this to create new variables

covid19_df <- covid19_date |> 
  mutate(Month = lubridate::month(Date, label = TRUE, abbr = FALSE)) |> 
  relocate(Month, .after = Date) |> 
  arrange(desc(Total))

covid19_df


```

<br />

------------------------------------------------------------------------

## Readr function: write_csv

the write_csv function allows us to export data frames to a csv file
once we are done cleaning it up or when we have done some analysis that
we want to export

```{r}
#| label: exporting data
#| eval: false

# now that we have this date frame cleaned let's save it

covid19_df <- covid19_month

covid19_df

# let's export the file

write_csv(covid19_df, "covid19_time_series.csv")

```

<br />

------------------------------------------------------------------------

## Haven function: Export as proprietary file

we can even export files that we have been working on as proprietary
files to work on in SPSS, SAS, or STATA

```{r}
#| label: export SPSS file
#| eval: false

# export to SPSS

haven::write_sav(covid19_df, "covid19_time_series.sav")

```

<br />

------------------------------------------------------------------------

# Part 4: Explore Your Data


-   [Psych Package](https://cran.r-project.org/web/packages/psych/index.html)

```{r}
#| label: Psych Package
#| eval: false

install.packages("psych")
library(psych)

```


```{r}
#| label: describe function
#| eval: false

describe(reviews_df)

```

-   [SummaryTools Package](https://cran.r-project.org/web/packages/summarytools/index.html)

```{r}
#| label: SummaryTools Package
#| eval: false

install.packages("summarytools")
library(summarytools)

```

```{r}
#| label: summarytools functions
#| eval: false

descr(reviews_df)

freq(reviews_df$Division_Name)

ctable(reviews_df$Recommended_IND, reviews_df$Division_Name)

```



-   [DataExplorer package](http://boxuancui.github.io/DataExplorer/)

```{r}
#| label: DataExplorer
#| eval: false

install.packages("DataExplorer")
library(DataExplorer)
```

```{r}
#| label: create report
#| eval: false

create_report(reviews_df)

```
